Timid tides or tall tails? \\ Investigating the outer reaches of the Sculptor and Ursa Minor dwarf galaxies

### 2025-05-02 thesis/introduction.tex

@@

\subsection{Sculptor}\label{sculptor}

\begin{figure}
\centering
\includegraphics[width=5.41667in,height=5.41667in]{figures/scl_des_dr2.png}
\caption[Picture of Sculptor]{Sculptor image from DES DR 2 via
HiPS2FITS.}\label{fig:scl_image}
\end{figure}

@@

Sculptor (Scl) is one of the first discovered dwarf galaxies of the
Milky Way (Shapley 1938; only preceded by the SMC and LMC!). As a
classical dwarf spheroidal, Scl is relatively bright and compact.

Since the initial discovery of Scl, most authors have noted that Scl has
a slight ellipticity (\(0.36\)), often attributed to tidal effects.
However, this does not align with the absolute proper motion or orbital
path of the galaxy.

While Scl has a relatively large pericentre (greater than 50kpc),
\citet{sestito+2023a} detect that the galaxy likely has an excess of
stars in the outskirts (past about 60 arcminutes). This excess is
perhaps one of the clearest indications that Scl may be affected by
tides. Here, our goal is to determine if under a \(\Lambda\) CMD
paradigm with DM-only simulations, tidal effects of the Milky Way (or
LMC) may indeed be consistent with these observations, or if these
observations may reveal instead a extended stellar ``halo'' or second
component of the galaxy---illustrating a complex formation for the
galaxy.

Theoretical work on Sculptor

\begin{itemize}
\tightlist
\item
  \citet{battaglia+2008}
\item
  \citet{iorio+2019}
\item
  \citet{amorisco+zavala+deboer2014}
\item
  \citet{battaglia+2008}
\item
  \citet{breddels+2013}
\item
  \citet{breddels+helmi2013}
\item
  \citet{richardson+fairbairn2014}
\item
  \citet{SFW2017}
\item
  \citet{innanen+papp1979}
\end{itemize}

Observational work on Scl

\begin{itemize}
\tightlist
\item
  \citet{sestito+2023a}
\item
  \citet{tolstoy+2023}, \citet{arroyo-polonio+2023},
  \citet{arroyo-polonio+2024}
\item
  \citet{eskridge1988}, \citet{eskridge1988a}, \citet{eskridge1988b}
\item
  \citet{coleman+dacosta+bland-hawthorn2005}
\item
  \citet{DQ1994}
\item
  \citet{WMO2009}
\item
  \citet{IH1995}, \citet{munoz+2018}
\item
  \citet{kirby+2009}
\item
  \citet{martinez-vazquez+2015}, \citet{pietrzynski+2008}
\end{itemize}

\begin{table*}[t]
\centering
\caption{Observed properties of Sculptor. References are: 1. Muñoz et al. (2018), 2. Tran et al. (2022), 3. (MV2020b?), 4 McConnachie and Venn (2020).}
\label{tbl:Observed-properties-of-Sculptor-References-are:-1-Muñoz-et-al-2018-2-Tran-et-al-2022-3-MV2020b?-4-McConnachie-and-Venn-2020}
\begin{tabular}{lll}
\toprule
parameter & value & Source\\
\midrule
$\alpha$ & $15.0183 \pm 0.0012^\circ$ & 1\\
$\delta$ & $-33.7186 \pm 0.0007^\circ$ & 1\\
distance modulus & $19.60 \pm 0.05$ (RRL) & 2\\
distance & $83.2 \pm 2$ kpc & 2\\
$\mu_{\alpha*}$ & $0.099 \pm 0.002 \pm 0.017$ mas yr$^{-1}$ & 3\\
$\mu_\delta$ & $-0.160 \pm 0.002_{\rm stat} \pm 0.017_{\rm sys}$ mas yr$^{-1}$ & 3\\
radial velocity & $111.2 \pm 0.2\ {\rm km\,s^{-1}}$ & this work\\
$\sigma_v$ & $9.67\pm0.16\ {\rm km\,s^{-1}}$ & this work\\
$R_h$ & $12.33 \pm 0.05$ arcmin & 4\\
$R_{h,inner}$ & $8.69 \pm 0.2$ arcmin & this work?\\
ellipticity & $0.36 \pm 0.01$ & 1\\
position angle & $92\pm1^\circ$ & 1\\
$M_V$ & $-10.82\pm0.14$ & 1\\
\bottomrule
\end{tabular}
\end{table*}

\subsection{Ursa Minor}\label{ursa-minor}

\begin{itemize}
\item
  \citet{sestito+2023b}
\item
  \citet{pace+2020}
\item
  \citet{bellazzini+2002}
\item
  \citet{hargreaves+1994}
\item
  \citet{martinez-delgado+2001}
\item
  \citet{munoz+2005}
\item
  \citet{pace+2020}
\item
  \citet{palma+2003}
\item
  \citet{spencer+2018}
\item
  \citet{vitral+2023}
\end{itemize}

\begin{figure}
\centering
\includegraphics[width=5.41667in,height=5.41667in]{figures/umi_image.jpg}
\caption[Picture of Ursa Minor]{Ursa Minor Dwarf. UMi appears as a
diagonal/elliptical haze of stars from the top left to the bottom right.
Even as as classical dwarf, Ursa minor is fairly diffuse and does not
stand obviously out from the background. Composite Public data/Giuseppe
Donatiello. Updated image in 2022.
https://www.flickr.com/photos/133259498@N05/page8}\label{fig:umi_image}
\end{figure}

\begin{table*}[t]
\centering
\caption{Observed properties of Ursa Minor. References are: (1) Muñoz et al. (2018), (2) Garofalo et al. (2025), (3) McConnachie and Venn (2020), (4) average of Pace et al. (2020) and Spencer et al. (2018).}
\label{tbl:Observed-properties-of-Ursa-Minor-References-are:-1-Muñoz-et-al-2018-2-Garofalo-et-al-2025-3-McConnachie-and-Venn-2020-4-average-of-Pace-et-al-2020-and-Spencer-et-al-2018}
\begin{tabular}{lll}
\toprule
parameter & value & Source\\
\midrule
$\alpha$ & $ 227.2420 \pm 0.0045$˚ & 1\\
$\delta$ & $67.2221 \pm 0.0016$˚ & 1\\
distance modulus & $19.23 \pm 0.11$ (RRL) & 2\\
distance & $70.1 \pm 3.6$ kpc & -\\
$\mu_\alpha*$ & $-0.124 \pm 0.004 \pm 0.017$ mas yr$^{-1}$ & 3\\
$\mu_\delta$ & $0.078 \pm 0.004_{\rm stat} \pm 0.017_{\rm sys}$ mas yr$^{-1}$ & 3\\
radial velocity & $-245.9 \pm 0.3_{\rm stat} \pm 1_{\rm sys}$ km s$^{-1}$ & this work\\
$\sigma_v$ & $8.8 \pm 0.2$ & this work\\
$r_h$ & $18.2 \pm 0.1$ arcmin & 1\\
ellipticity & $0.55 \pm 0.01$ & 1\\
position angle & $50 \pm 1^\circ$ & 1\\
$M_V$ & $-9.03 \pm 0.05$ & 1\\
\bottomrule
\end{tabular}
\end{table*}

@@

An emperical estimate of where the simulation's stars are becoming
unbound is, as stated in \citet{PNM2008}, the break radius \[
R_b = C\,\sigma_{v}\,\Delta t
\] where \(\sigma_v\) is the present line of sight velocity dispersion ,
\(\delta t\) is the time since pericentre, and \(C \approx 0.55\) is a
fit. The idea motivating this equation is stars in the inner regions
will have dynamically equilibriated to the new potential (phase mixed),
however the outer regions are no longer in steady state, so we have to
wait until the crossing time reaches them as well.

@@

From the \citet{nfw1996} paper, eqns. 3 \& 4 \[
\frac{\rho}{\rho_c} = \frac{\delta_c}{(r/r_s)(1+r/r_s)^2}
\] where \[
\delta_c = \frac{200}{3}\frac{c^3}{[\ln(1+c)-c/(1+c)]}
\] \(c\) is concentration parameter, \(\rho_c\) is the critical density
of the universe, and \(r_s\) is the characteristic scale length of the
halo.

The NFW halo is sometimes described by \(M_{200}\). \(r_{200}\) is the
radius at which the mean density of the halo interior to \(r_{200}\) is
200 times the critical density of the universe, and \(M_{200}\) is the
mass contained inside \(r_{200}\). As equations: \[
\rho_{200} = 200\rho_{c} = \frac{M_{200}}{(4\pi/3) r_{200}^3}
\]

\[
r_{200} = \sqrt[3]{\frac{1}{200}\frac{3M_{200}}{4\pi \rho_{\rm c
}}}
\]

\[
M_{200} = \frac{4\pi}{3} r_{200}^3\ \rho_{200}
\]

\(M_{200}\) is also sometimes called the virial mass of the halo.
\(r_{200}\) is directly related to \(r_s\) by \[
r_{200} = c\,r_s
\]

Another useful definition is \[
A(x) \equiv \log (1+x) - \frac{x}{1+x}.
\]

We will also define a dimensionless radius \[
x \equiv r/r_s.
\]

@@

\[
\rho(x) =  \frac{\rho_s/3}{x\ (1+x)^2}
\] where \[
\rho_s = \frac{c^3}{A(c)} \rho_{200}
\] and \(A(c)\) is as above. The characteristic density can also be
written in terms of scale mass, \(M_s = M_{200}/{A(c)}\) (see below),
giving \[
\rho_s = \frac{c^3}{A(c)} \frac{M_{200}}{(4\pi/3)\ r_{200}^3}  = \frac{3M_s}{4\pi\, {r_s}^3}
\] Note that the NFW density profile is the same as an alpha-beta-gamma
profile where \(\alpha=\gamma=1\) and \(\beta =3\).

@@

The circular velocity in terms of
\(v_{200} = \sqrt{G M_{200} / R_{200}}\) is \[
\left(v_{\rm circ}/v_{200}\right)^2 = \frac{A(x)/x}{A(c)/c},
\]

or in terms of \(M_s\) and \(r_s\), \[
v_{\rm circ}^2 = \frac{G M(r)}{r} = \frac{G M_s A(r/r_s)}{r}.
\] Another parameterization of the NFW profile is in terms of the
maximum circular velocity \(v_{\rm circ}^{\rm max}\) and the radius at
which it is reached \(r_{\rm circ}^{\rm max}\). Given the scale radius,
\[
r_{\rm circ}^{\rm max} = \alpha\ r_s
\]

>>>>

\begin{itemize}
\tightlist
\item
  \citet{fattahi+2013}, \citet{fattahi+2018}
\item
  \citet{sanchez-salcedo+hernandez2007}: mond in dsph
\item
  \citet{mayer+2001} theory of tidal stripping
\item
  \citet{IH1995} structural parameters
\end{itemize}

@@

\includegraphics[width=5.41667in,height=5.41667in]{/Users/daniel/thesis/figures/scl_des_dr2.png}Figure:
Image of the Sculptor dwarf spheroidal galaxy from Dark Energy Survey
Data Release 2 {[}\citet{abbott+2021}; image created with HiPS2FITS{]}.
Sculptor appears as a fairly prominent, extended over density of
predominantly faint, red stars. 0.5 degree field of view centred on
Sculptor.

\includegraphics[width=5.41667in,height=5.41667in]{figures/umi_DSS2_0.75deg.png}Figure:
Image of Ursa Minor dwarf galaxy from the Digitized Sky Survey 2 (0.75
deg field of view tangent plane). UMi appears as a diagonal/elliptical
haze of faint, reddish stars from the top left to the bottom right. Even
as classical dwarf, Ursa minor is fairly diffuse and does not stand
obviously out from the background.

@@

@@

An emperical estimate of where the simulation's stars are becoming
unbound is, as stated in \citet{PNM2008}, the break radius
\begin{equation}
R_b = C\,\sigma_{v}\,\Delta t
\end{equation} where \(\sigma_v\) is the present line of sight velocity
dispersion , \(\delta t\) is the time since pericentre, and
\(C \approx 0.55\) is a fit. The idea motivating this equation is stars
in the inner regions will have dynamically equilibriated to the new
potential (phase mixed), however the outer regions are no longer in
steady state, so we have to wait until the crossing time reaches them as
well.

@@

From the \citet{nfw1996} paper, eqns. 3 \& 4 \begin{equation}
\frac{\rho}{\rho_c} = \frac{\delta_c}{(r/r_s)(1+r/r_s)^2}
\end{equation} where \begin{equation}
\delta_c = \frac{200}{3}\frac{c^3}{[\ln(1+c)-c/(1+c)]}
\end{equation} \(c\) is concentration parameter, \(\rho_c\) is the
critical density of the universe, and \(r_s\) is the characteristic
scale length of the halo.

The NFW halo is sometimes described by \(M_{200}\). \(r_{200}\) is the
radius at which the mean density of the halo interior to \(r_{200}\) is
200 times the critical density of the universe, and \(M_{200}\) is the
mass contained inside \(r_{200}\). As equations: \begin{equation}
\rho_{200} = 200\rho_{c} = \frac{M_{200}}{(4\pi/3) r_{200}^3}
\end{equation}

\begin{equation}
r_{200} = \sqrt[3]{\frac{1}{200}\frac{3M_{200}}{4\pi \rho_{\rm c
}}}
\end{equation}

\begin{equation}
M_{200} = \frac{4\pi}{3} r_{200}^3\ \rho_{200}
\end{equation}

\(M_{200}\) is also sometimes called the virial mass of the halo.
\(r_{200}\) is directly related to \(r_s\) by \begin{equation}
r_{200} = c\,r_s
\end{equation}

Another useful definition is \begin{equation}
A(x) \equiv \log (1+x) - \frac{x}{1+x}.
\end{equation}

We will also define a dimensionless radius \begin{equation}
x \equiv r/r_s.
\end{equation}

@@

\begin{equation}
\rho(x) =  \frac{\rho_s/3}{x\ (1+x)^2}
\end{equation} where \begin{equation}
\rho_s = \frac{c^3}{A(c)} \rho_{200}
\end{equation} and \(A(c)\) is as above. The characteristic density can
also be written in terms of scale mass, \(M_s = M_{200}/{A(c)}\) (see
below), giving \begin{equation}
\rho_s = \frac{c^3}{A(c)} \frac{M_{200}}{(4\pi/3)\ r_{200}^3}  = \frac{3M_s}{4\pi\, {r_s}^3}
\end{equation} Note that the NFW density profile is the same as an
alpha-beta-gamma profile where \(\alpha=\gamma=1\) and \(\beta =3\).

@@

The circular velocity in terms of
\(v_{200} = \sqrt{G M_{200} / R_{200}}\) is \begin{equation}
\left(v_{\rm circ}/v_{200}\right)^2 = \frac{A(x)/x}{A(c)/c},
\end{equation}

or in terms of \(M_s\) and \(r_s\), \begin{equation}
v_{\rm circ}^2 = \frac{G M(r)}{r} = \frac{G M_s A(r/r_s)}{r}.
\end{equation} Another parameterization of the NFW profile is in terms
of the maximum circular velocity \(v_{\rm circ}^{\rm max}\) and the
radius at which it is reached \(r_{\rm circ}^{\rm max}\). Given the
scale radius, \begin{equation}
r_{\rm circ}^{\rm max} = \alpha\ r_s
\end{equation}

### 2025-05-02 thesis/methods.tex

\[
\Phi(r) = - \frac{GM}{r + a}
\] where \(a=1.3\,{\rm kpc}\) is the scale radius and
\(M=2.1 \times 10^{10}\,\Mo\) is the total mass.

The thin and thick disks are represented with the
\citet{miyamoto+nagai1975} cylindrical potential: \[
\Phi(R, z) = \frac{-GM}{\left(R^2 + \left[a + \sqrt{z^2 + b^2}\right]^{2}\right)^{1/2}}
\]

@@

We adopt a softening length of \[
h_{\rm grav} = 0.014 \left(\frac{r_s}{2.76\,{\rm kpc}}\right)\left(\frac{N}{10^7}\right)^{-1/2}.
\] See appendix REF for a discussion of this choice, which is similar to
the \citet{power+2003} suggested softening.

@@

As discussed in \citet{power+2003}, the region where density is
converged is related to the region which becomes collisionally relaxed
over the time of the universe (so about 5-10 Gyr). (i.e.~where the
collisionless assumption breaks down). The relaxation timescale is given
by \(t_{\rm relax} = t_{\rm circ} N(r) / 8\ln N(r)\) or otherwise, \[
t_{\rm relax}(r) := t_{\rm circ}(r) \frac{N(r)}{8\,\ln N(r)}
= {t_{\rm circ}(r_{200})} \frac{\sqrt{200}}{8} \frac{N(r)}{\ln N(r)} \left(\frac{\bar \rho (r)}{\rho_{\textrm crit}}\right)^{-1/2}
\] This works out to be about 6-10 times (increasing with particle
number) our adopted softening length for NFW halos given our
assumptions. As such, at full resolution, we can only trust density
profiles down to \(\sim10\epsilon\), just enough to resolve stellar
density profiles.

@@

\[
v_{\rm circ}(r) = \sqrt{\frac{GM(r)}{r}}
\]

@@

Let \(\Psi\) be the potential (normalized to vanish at infinity) and
\({\cal E}\) is the binding energy \({\cal E} = \Psi - 1/2 v^2\). If we
know \(f({\cal E})\), the distribution function (phase-space density in
energy), then we assign the stellar weight for a given particle with
energy \({\cal E}\) is \[
P_\star({\cal E}) = \frac{f_\star({\cal E})}{f_{\rm halo}({\cal E})}.
\] While \(f({\cal E})\) is a phase-space density, the differential
energy distribution includes an additional \(g({\cal E})\) occupation
term (BTXXX).

We use Eddington inversion to find the distribution function, (eq.
4-140b in BT87) \[
f({\cal E}) = \frac{1}{\sqrt{8}\, \pi^2}\left( \int_0^{\cal E} \frac{d^2\rho}{d\Psi^2} \frac{1}{\sqrt{{\cal E} - \Psi}}\ d\Psi + \frac{1}{\sqrt{\cal E}} \left(\frac{d\rho}{d\Psi}\right)_{\Psi=0} \right).
\]

@@

The 2D exponential stellar profile is given by \[
\Sigma_\star(R) = A e^{-R / R_s}
\] where \(A\) is some normalization and \(R_s\) is the exponential
scale radius. The 3D deprojected profile is found through abel inversion
(e.g.~Rapha\ldots), i.e. \[
\rho_\star (r) =- \frac{1}{\pi}\int_r^\infty \frac{d\Sigma}{dR} \frac{1}{\sqrt{R^2 - r^2}} dR  = \frac{\Sigma_0}{\pi R_s^2}\,K_0(r/R_s)
\] where \(K_0\) is the 0th order modified Bessel function of the second
type.

@@

A Plummer profile is defined by \[
\Sigma(R) = \frac{M}{4\pi R_s^2} \left(1 + \frac{R^2}{R_s^2}\right)^{-2} ,
\] where \(M\) is the total mass and \(R_s\) is the characteristic scale
radius. The density is \[
\rho(r) = \frac{3M}{4\pi\,R_s^3} \left(1 + \frac{r^2}{R_s^2}\right)^{-5/2}.
\]

>>>>

\begin{equation}
\Phi(r) = - \frac{GM}{r + a}
\end{equation} where \(a=1.3\,{\rm kpc}\) is the scale radius and
\(M=2.1 \times 10^{10}\,\Mo\) is the total mass.

The thin and thick disks are represented with the
\citet{miyamoto+nagai1975} cylindrical potential: \begin{equation}
\Phi(R, z) = \frac{-GM}{\left(R^2 + \left[a + \sqrt{z^2 + b^2}\right]^{2}\right)^{1/2}}
\end{equation}

@@

We adopt a softening length of \begin{equation}
h_{\rm grav} = 0.014 \left(\frac{r_s}{2.76\,{\rm kpc}}\right)\left(\frac{N}{10^7}\right)^{-1/2}.
\end{equation} See appendix REF for a discussion of this choice, which
is similar to the \citet{power+2003} suggested softening.

@@

As discussed in \citet{power+2003}, the region where density is
converged is related to the region which becomes collisionally relaxed
over the time of the universe (so about 5-10 Gyr). (i.e.~where the
collisionless assumption breaks down). The relaxation timescale is given
by \(t_{\rm relax} = t_{\rm circ} N(r) / 8\ln N(r)\) or otherwise,
\begin{equation}
t_{\rm relax}(r) := t_{\rm circ}(r) \frac{N(r)}{8\,\ln N(r)}
= {t_{\rm circ}(r_{200})} \frac{\sqrt{200}}{8} \frac{N(r)}{\ln N(r)} \left(\frac{\bar \rho (r)}{\rho_{\textrm crit}}\right)^{-1/2}
\end{equation} This works out to be about 6-10 times (increasing with
particle number) our adopted softening length for NFW halos given our
assumptions. As such, at full resolution, we can only trust density
profiles down to \(\sim10\epsilon\), just enough to resolve stellar
density profiles.

@@

\begin{equation}
v_{\rm circ}(r) = \sqrt{\frac{GM(r)}{r}}
\end{equation}

@@

Let \(\Psi\) be the potential (normalized to vanish at infinity) and
\({\cal E}\) is the binding energy \({\cal E} = \Psi - 1/2 v^2\). If we
know \(f({\cal E})\), the distribution function (phase-space density in
energy), then we assign the stellar weight for a given particle with
energy \({\cal E}\) is \begin{equation}
P_\star({\cal E}) = \frac{f_\star({\cal E})}{f_{\rm halo}({\cal E})}.
\end{equation} While \(f({\cal E})\) is a phase-space density, the
differential energy distribution includes an additional \(g({\cal E})\)
occupation term (BTXXX).

We use Eddington inversion to find the distribution function, (eq.
4-140b in BT87) \begin{equation}
f({\cal E}) = \frac{1}{\sqrt{8}\, \pi^2}\left( \int_0^{\cal E} \frac{d^2\rho}{d\Psi^2} \frac{1}{\sqrt{{\cal E} - \Psi}}\ d\Psi + \frac{1}{\sqrt{\cal E}} \left(\frac{d\rho}{d\Psi}\right)_{\Psi=0} \right).
\end{equation}

@@

The 2D exponential stellar profile is given by \begin{equation}
\Sigma_\star(R) = A e^{-R / R_s}
\end{equation} where \(A\) is some normalization and \(R_s\) is the
exponential scale radius. The 3D deprojected profile is found through
abel inversion (e.g.~Rapha\ldots), i.e. \begin{equation}
\rho_\star (r) =- \frac{1}{\pi}\int_r^\infty \frac{d\Sigma}{dR} \frac{1}{\sqrt{R^2 - r^2}} dR  = \frac{\Sigma_0}{\pi R_s^2}\,K_0(r/R_s)
\end{equation} where \(K_0\) is the 0th order modified Bessel function
of the second type.

@@

A Plummer profile is defined by \begin{equation}
\Sigma(R) = \frac{M}{4\pi R_s^2} \left(1 + \frac{R^2}{R_s^2}\right)^{-2} ,
\end{equation} where \(M\) is the total mass and \(R_s\) is the
characteristic scale radius. The density is \begin{equation}
\rho(r) = \frac{3M}{4\pi\,R_s^3} \left(1 + \frac{r^2}{R_s^2}\right)^{-5/2}.
\end{equation}
